---
title: "Equity-Informed Access Improvement Simulator for British Columbia"
subtitle: "A reproducible, synthetic-data capability demonstrator (version 2.0.0)"
author: "Christopher Bwanika"
date: last-modified
date-format: "D MMMM YYYY"
---

```{r setup}
if (!file.exists(file.path("outputs", "run_manifest.csv"))) {
  stop("No outputs found. Run run_all.R (and tests/run_tests.R) before rendering.")
}
out <- function(name) utils::read.csv(file.path("outputs", name), check.names = FALSE)
fig <- function(name) knitr::include_graphics(file.path("outputs", "figures", name))
num <- function(x, d = 0) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x, d = 0) paste0(formatC(100 * x, format = "f", digits = d), "%")
tab <- function(df, digits = 3, ...) knitr::kable(df, digits = digits, row.names = FALSE,
                                                  format.args = list(big.mark = ","), ...)
lab <- c(usual_care = "Usual care", targeted_navigation = "Targeted navigation",
         universal_navigation = "Universal navigation", lower_barrier = "Lower barrier",
         moderate_barrier = "Moderate barrier", higher_barrier = "Higher barrier", all = "All")
relabel <- function(x) ifelse(x %in% names(lab), lab[x], x)
options(knitr.kable.NA = "")

pars <- utils::read.csv(file.path("data", "parameters.csv"))
pv <- function(p, g = "all") pars$value[pars$parameter == p & pars$group == g]
k0 <- pv("k_threshold")
bs <- out("base_case_by_strategy.csv"); inc <- out("base_case_incremental.csv")
vs <- out("base_case_vs_usual_care.csv"); grp <- out("base_case_by_group.csv")
ceac <- out("psa_ceac_evpi.csv"); p_at_k <- ceac[ceac$k == k0, ]
icer_t <- vs$icer_vs_usual_care[vs$strategy == "targeted_navigation"]
icer_ut <- inc$icer[inc$strategy == "universal_navigation"]
opt <- bs$strategy[bs$optimal_at_k]
grp_icer <- sapply(c("lower_barrier", "moderate_barrier", "higher_barrier"), function(g) {
  a <- grp[grp$strategy == "universal_navigation" & grp$group == g, ]
  b <- grp[grp$strategy == "usual_care" & grp$group == g, ]
  (a$cost - b$cost) / (a$qaly - b$qaly)
})
dsum <- out("dcea_psa_summary.csv"); trade <- out("dcea_tradeoff.csv")
cross <- trade$epsilon[trade$k == 100000 & trade$preferred == "targeted_navigation"]
cross_eps <- if (length(cross) > 0) min(cross) else NA
```

::: {.callout-warning}
## Synthetic-data capability demonstrator
All data, inputs and results in this report are **simulated** to illustrate methods. This is
not an official analysis, recommendation, endorsement or evaluation by any government, health
authority, agency or community, and no result describes a real population or programme.
:::

# Summary

**Decision question (hypothetical).** For adults referred into a diagnostic or specialist
pathway, should a health system offer patient navigation to everyone, only to patients facing
the highest structural barriers, or to no one (usual care)?

**What the prototype shows (illustrative results only).**

- In the base case, targeted navigation costs CAD `r num(icer_t)` per QALY gained compared
  with usual care, and moving from targeted to universal navigation costs CAD
  `r num(icer_ut)` per QALY. At CAD `r num(k0)` per QALY the strategy with the highest net
  monetary benefit is **`r relabel(opt)`**.
- Navigation is far better value where barriers are highest: CAD
  `r num(grp_icer[["higher_barrier"]])` per QALY in the higher-barrier group against CAD
  `r num(grp_icer[["lower_barrier"]])` in the lower-barrier group.
- In the probabilistic analysis, the probability that targeted navigation is optimal at CAD
  `r num(k0)` per QALY is `r pct(p_at_k$prob_optimal[p_at_k$strategy == "targeted_navigation"])`
  (universal: `r pct(p_at_k$prob_optimal[p_at_k$strategy == "universal_navigation"])`).
- Distributional analysis: targeted navigation both adds health and narrows the gap between
  groups in `r pct(dsum$prob_win_win[dsum$strategy == "targeted_navigation"], 1)` of PSA
  draws. At CAD 100,000 per QALY universal navigation yields more total health, and targeted
  navigation is preferred only with Atkinson inequality aversion of at least
  `r cross_eps` - a genuine equity-efficiency trade-off.

These are **assumption-dependent simulated outputs**, produced to demonstrate an auditable
workflow. They are not evidence about any real programme.

# Decision problem

| Element | Specification |
|---|---|
| Population | Synthetic cohort of adults aged 55 referred into a diagnostic/specialist pathway, in three structural-barrier groups |
| Strategies | Usual care; targeted navigation (higher-barrier group only); universal navigation (all groups) |
| Perspective | Publicly funded health system (costs); household non-medical costs reported separately (ECEA-lite) |
| Time horizon | Lifetime (to age 100), quarterly cycles |
| Discounting | 1.5% per year for costs and QALYs (CDA-AMC reference case); 0% and 3% in scenarios |
| Outcomes | Costs, QALYs, life-years, acute events, time unresolved; NMB; distribution of health; household burden |
| Currency | Canadian dollars, illustrative values (no price-year adjustment applies to synthetic inputs) |
| Threshold | CAD `r num(k0)` per QALY for illustration; results shown from 0 to 150,000 |

# Model

A cohort state-transition (Markov) model with four states and quarterly cycles. Transitions
are built as sequential conditional probabilities, so every row of every transition matrix
sums to one by construction. Navigation multiplies the hazard of completing the pathway for
navigated patients while support lasts (`r pv("nav_cycles")` quarters); uptake is lower in
higher-barrier groups. Offered groups are modelled as an exact uptake-weighted mix of a
navigated and a non-navigated sub-cohort.

```{mermaid}
flowchart LR
  U[Unresolved] -->|pathway completed| M[Managed]
  M -->|disengages| U
  U -->|acute event| A[Acute event<br/>one quarter]
  M -->|acute event| A
  A -->|survives, enters care| M
  A -->|survives, not in care| U
  U --> D[Dead]
  M --> D
  A --> D
```

```{r fig-trace, fig.cap="Cohort trace for the higher-barrier group, first 10 years."}
fig("fig01_trace.png")
```

## Inputs

Every input, its distribution and its rationale is held in `data/parameters.csv`.
`value` is the distribution mean and `se` its standard error.

```{r inputs}
p <- pars[pars$module %in% c("markov", "ecea"),
          c("parameter", "group", "value", "se", "distribution", "unit", "source_type")]
tab(p, digits = 4)
```

# Cost-effectiveness results

## Base case (deterministic)

```{r base}
b <- bs[, c("strategy", "cost", "qaly", "ly", "acute_events", "nmb")]
b$strategy <- relabel(b$strategy)
tab(b, digits = 4, col.names = c("Strategy", "Cost (CAD)", "QALYs", "Life-years",
                                 "Acute events", "NMB (CAD)"))
```

```{r frontier}
f <- inc[, c("strategy", "status", "inc_cost", "inc_effect", "icer")]
f$strategy <- relabel(f$strategy)
tab(f, digits = 4, col.names = c("Strategy", "Status", "Incremental cost", "Incremental QALYs",
                                 "ICER (CAD/QALY)"))
```

Per-patient results by group (offered groups include uptake):

```{r by-group}
g <- grp[, c("strategy", "group", "cost", "qaly", "acute_events", "time_unresolved", "managed_1y")]
g$strategy <- relabel(g$strategy); g$group <- relabel(g$group)
tab(g, digits = 4)
```

## One-way and scenario analyses

```{r fig-tornado, fig.cap="One-way sensitivity analysis: each input at its 2.5th and 97.5th PSA percentile."}
fig("fig05_tornado.png")
```

```{r scenarios}
s <- out("scenarios.csv")
s$strategy <- relabel(s$strategy); s$optimal_strategy <- relabel(s$optimal_strategy)
tab(s, digits = 4)
```

## Probabilistic sensitivity analysis

`r num(nrow(out("psa_parameter_draws.csv")))` iterations; one parameter draw is shared by all
strategies and groups within an iteration.

```{r psa-summary}
ps <- out("psa_summary.csv"); ps$strategy <- relabel(ps$strategy)
tab(ps[, c("strategy", "mean_inc_cost", "inc_cost_lo", "inc_cost_hi",
           "mean_inc_qaly", "inc_qaly_lo", "inc_qaly_hi")], digits = 4,
    col.names = c("Strategy", "Mean incremental cost", "2.5%", "97.5%",
                  "Mean incremental QALYs", "2.5%", "97.5%"))
```

```{r fig-ce-plane, fig.cap="Cost-effectiveness plane."}
fig("fig02_ce_plane.png")
```

```{r fig-ceac, fig.cap="Cost-effectiveness acceptability curves and frontier."}
fig("fig03_ceac.png")
```

```{r fig-evpi, fig.cap="Expected value of perfect information for one annual cohort."}
fig("fig04_evpi.png")
```

```{r fig-conv, fig.cap="Convergence of the PSA mean incremental NMB."}
fig("fig06_convergence.png")
```

# Distributional cost-effectiveness analysis

Health gains by group are combined with health opportunity costs (incremental cost divided by
the threshold), shared across groups by an explicit rule. The base case shares them in
proportion to population; two other rules are reported as scenarios, informed by English
evidence that displaced health falls more heavily on deprived groups.

```{r dcea-metrics}
m <- out("dcea_metrics.csv")
m <- m[, c("strategy", "oc_scenario", "total_nhb", "gap_reduction_days",
           "ew_nhb_atkinson_0", "ew_nhb_atkinson_5", "ew_nhb_atkinson_10", "ew_nhb_kolm_0.15")]
m$strategy <- relabel(m$strategy)
tab(m, digits = 3, col.names = c("Strategy", "Opportunity-cost rule", "Net health benefit (QALYs)",
                                 "Gap reduction (days/person)", "EW-NHB e=0", "EW-NHB e=5",
                                 "EW-NHB e=10", "EW-NHB Kolm 0.15"))
```

```{r fig-dcea-dist, fig.cap="Net health change per person by group."}
fig("fig07_dcea_distribution.png")
```

```{r fig-equity-plane, fig.cap="Equity-efficiency impact plane (PSA draws and base case)."}
fig("fig08_equity_plane.png")
```

```{r fig-ede, fig.cap="Equity-weighted net health benefit across inequality aversion."}
fig("fig09_ede_epsilon.png")
```

```{r dcea-psa}
d <- dsum[, c("strategy", "prob_health_gain", "prob_inequality_reduced", "prob_win_win")]
d$strategy <- relabel(d$strategy)
tab(d, digits = 3, col.names = c("Strategy", "P(more health)", "P(smaller gap)", "P(both)"))
```

Inequality aversion is a value judgement, so results are shown across a range. For reference,
a survey of the English public found median values of about 11 (Atkinson) and 0.15 (Kolm).

# ECEA-lite: household non-medical burden

First-year household non-medical costs (travel, time off work, missed appointments, lost income
during acute events) are simulated for individual patients through the same transition
matrices as the cohort model. This is a financial-burden extension, not a full ECEA and not a
measure of catastrophic health expenditure.

```{r ecea}
e <- out("ecea_results.csv")
e <- e[, c("strategy", "group", "mean_hh_cost", "share_above_0.025", "share_above_0.05",
           "share_above_0.1", "high_burden_averted_per_1000")]
e$strategy <- relabel(e$strategy); e$group <- relabel(e$group)
tab(e, digits = 4)
```

```{r fig-ecea, fig.cap="Share of households above the high-burden threshold."}
fig("fig10_ecea.png")
```

Cross-validation of the microsimulation against the cohort model:

```{r ecea-check}
tab(out("ecea_cross_check.csv"), digits = 2)
```

# Causal-evidence demonstrations

These modules use synthetic data with **known, planted effects**. Their purpose is to show that
the estimation pipelines recover what was planted and pass standard falsification checks, not
to estimate any real effect.

## Interrupted time series

```{r fig-its, fig.cap="Synthetic ITS: observed, fitted and counterfactual series."}
fig("fig11_its.png")
```

```{r its-est}
ie <- out("its_estimates.csv"); ie$group <- relabel(ie$group)
tab(ie[, c("group", "analysis", "term", "estimate", "ci_lo", "ci_hi", "planted")], digits = 4)
```

```{r its-cal}
ic <- out("its_calibration.csv"); ic$group <- relabel(ic$group)
tab(ic, digits = 3, col.names = c("Group", "Mean bias", "95% CI coverage", "Placebo false-positive rate"))
```

The calibration re-generates the synthetic data many times. With REML estimation, coverage is
close to the nominal 95% and the placebo test's false-positive rate is close to 5%; maximum
likelihood gave noticeably narrower, over-confident intervals in development, consistent with
published simulation evidence on short interrupted time series.

Bridge from the ITS estimate to the Markov input (a consistency check, not new evidence):

```{r its-bridge}
ib <- out("its_bridge.csv"); ib$group <- relabel(ib$group)
tab(ib, digits = 3, col.names = c("Group", "Model input HR", "ITS-implied HR", "2.5%", "97.5%"))
```

## Regression discontinuity

```{r fig-rdd, fig.cap="Synthetic RDD: 90-day completion around the eligibility cutoff."}
fig("fig12_rdd.png")
```

```{r rdd-est}
tab(rbind(out("rdd_estimates.csv"), out("rdd_checks.csv"))[, c("analysis", "estimate",
    "robust_ci_lo", "robust_ci_hi", "bandwidth", "planted", "covers_planted")], digits = 3)
```

Density (manipulation) test p-value: `r num(out("rdd_density_test.csv")$p_value, 3)`.

```{r rdd-cal}
tab(out("rdd_calibration.csv"), digits = 3, col.names = c("Estimator", "Mean bias", "95% CI coverage"))
```

# Discrete event simulation: clinic capacity and no-shows

A booked slot is used whether or not the patient attends, so no-shows waste scarce capacity.
Because navigation reduces no-shows, it can shorten waits for everyone - including patients who
are not navigated - a system-level effect a cohort model cannot represent.

```{r fig-des-queue, fig.cap="Diagnostic waitlist over time."}
fig("fig13_des_queue.png")
```

```{r fig-des-waits, fig.cap="Days from referral to completed appointment."}
fig("fig14_des_waits.png")
```

```{r des}
ds <- out("des_summary.csv"); ds$strategy <- relabel(ds$strategy); ds$group <- relabel(ds$group)
tab(ds[order(ds$group, ds$strategy), ], digits = 3)
```

# Verification and validation

```{r tests, results="asis"}
if (file.exists(file.path("outputs", "test_results.csv"))) {
  tr <- out("test_results.csv")
  cat(sprintf("The automated test suite ran **%d tests** (%s expectations passed; %d failed or errored).",
              nrow(tr), num(sum(tr$passed)), sum(tr$failed) + sum(tr$error)), "\n\n")
  tr$one <- 1
  print(knitr::kable(stats::aggregate(cbind(tests = one, expectations = passed, failed) ~ file,
                                      data = tr, FUN = sum), row.names = FALSE))
} else {
  cat("Test results not found: run `tests/run_tests.R` before rendering.\n")
}
```

The tests cover every defect found in the v1 blueprint, known-answer checks of the engine,
null and direction tests, distribution properties, accounting identities in the DCEA, the
microsimulation-cohort cross-check, planted-effect recovery in the causal modules, and
regression tests against stored reference values. See `methods/verification.md`.

# Limitations

- Every input is synthetic. No result is calibrated to, or validated against, real data.
- The cohort model is a single-age cohort; age, comorbidity and repeat referrals are not modelled.
- Parameters are sampled independently; correlations between inputs are not represented.
- The navigation effect is assumed to act only on pathway completion and only while support lasts.
- Opportunity-cost distribution rules and inequality aversion are value-laden scenario choices.
- The ITS and RDD data are generated from the model's own assumptions, so their agreement with
  the model is a consistency check, not independent evidence.
- The DES represents one stylised clinic and is not linked to the cohort model's outcomes.
- Barrier groups are neutral constructs; any real application needs community co-design,
  data governance and ethical review before groups are defined.

# Reporting checklist (CHEERS 2022)

| # | Item | Where addressed |
|---|---|---|
| 1-2 | Title; abstract | Title; Summary |
| 3 | Background and objectives | Summary; Decision problem |
| 4 | Health economic analysis plan | `methods/protocol.md` |
| 5-10 | Population, setting, comparators, perspective, time horizon, discount rate | Decision problem |
| 11-13 | Selection, measurement and valuation of outcomes | Model; Inputs |
| 14-15 | Resources and costs; currency and price date | Inputs; Decision problem |
| 16-17 | Rationale and description of model; analytics and assumptions | Model; `methods/model_specification.md` |
| 18 | Characterising heterogeneity | Results by group |
| 19 | Characterising distributional effects | DCEA; ECEA-lite |
| 20 | Characterising uncertainty | One-way, scenario and probabilistic analyses |
| 21 | Approach to engagement with patients and others | Not undertaken (synthetic demonstrator); required before real use |
| 22-24 | Study parameters; main results; effect of uncertainty | Inputs; Results |
| 25 | Effect of engagement | Not applicable (no engagement undertaken) |
| 26 | Findings, limitations, generalisability | Summary; Limitations |
| 27-28 | Funding; conflicts of interest | Unfunded; none declared |

# Reproducibility

Run `run_all.R`, then `tests/run_tests.R`, then `quarto render`. The package versions are
pinned in `renv.lock`; every output's MD5 checksum is recorded in `outputs/run_manifest.csv`.

```{r manifest}
mf <- out("run_manifest.csv")
tab(mf[mf$section %in% c("run", "package"), c("section", "item", "value")])
```

`r sum(mf$section == "file_md5")` output files were checksummed in this run.
